---
title: "The Progress Report I"
author: "Dereck J, Caileigh L, Ben H"
date: "3/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Description
Caileigh Lavigne, Dereck Jernigan, and Ben Hudgens, will be writing a program that will monitor the pH of a reaction, while calculating the pKa, Ka, and the concentration of the unknown solution. 

## Dereck J:

I have added a table to R Markdown, eventually the coded equations will be inserted into the table for organization.

## Caileigh Lavigne
I have located the pH meter that will be used for this project. I have also began looking at material that will help me better understand how to communicate with the pH meter, as well as receive its output data.

```{r Table}
Table <- read.table(header=TRUE, text="
pH    pKa   Ka    Concentration
0     0     0       0
")

```

## Websites Consulted
https://files.atlas-scientific.com/pH_EZO_Datasheet.pdf
https://www.edn.com/how-to-automate-measurements-with-python/
https://www.adamsmith.haus/python/answers/how-to-create-a-timed-loop-in-python
